The Galactic Chemical Evolution of Carbon: Implications for Stellar Nucleosynthesis

### 2025-03-06 carbon_paper/tex/main.tex
updated draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,
%referee, % this line changes to double-spaced 1 column
usenatbib]{mnras}

@@

\newcommand{\fruity}{\texttt{\hyperlink{fruity}{Fruity}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{1.44}
\newcommand{\nsubgiants}{14,069}

@@

% other
\newcommand{\ycmg}{\ensuremath{2.7 + 32\left(Z-\Zo\right)}}
\newcommand{\fmeas}{20\%}

@@

\newcommand{\Yct}{{y_{\rm C}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\ycagb}{\y_{\rm C}^{\rm AGB}}
\newcommand{\aagb}{\alpha_{\rm C}^{\rm AGB}}
\newcommand{\zagb}{\zeta_{\rm C}^{\rm AGB}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}
\newcommand{\noneqagb}{\nu_{\rm C}^{\rm AGB}}
 \newcommand{\yb}{\ensuremath{\rotatebox[origin=B,y=0.5ex]{180}{y}}}
\newcommand{\y}{Y}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}

\newcommand{\zetao}{\zeta^{(0)}}
\newcommand{\zetai}{\zeta^{(1)}}
\newcommand{\zetaii}{\zeta^{(2)}}

@@

% Abstract of the paper
\begin{abstract}
% context
The astrophysical origin of carbon remains largely uncertain---different studies propose that either asymptotic giant branch (AGB) stars or core collapse supernovae (CCSNe) may be the dominant process.
% 
We aim to constrain the stellar yields of C through multi-zone Galactic chemical evolution models by comparing predictions with \apogee\ subgiants abundances.
% 
We find that \caafe\ trends, when restricted in metallicity, are an empirical estimate of the delayed C sources, enabling us to estimate that \agb\ stars and \cc\ produce about 20\% and 80\% of C, respectively.  
The \caah\ trend instead represents the equilibrium abundances of C and Mg. 
We use the \caah\ trend to estimate total C/Mg yield ratio, determining that  $\Ycc/\Yoc = EQUATION$.
% misc
Our models are relatively independent of uniform scaling of yields and outflows, and alternate star formation histories. 
However, the stars which contribute to \agb\ C production and the \ia\ delay time distribution of Fe contribute uncertainties to our conclusions. 
% gas phase
While reliable gas-phase and low-metallicity measurements of C are challenging, we find that our model and a single-zone model with our recommended yields replicate the broad trends of \caah{} across different environments and metallicities.

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio against [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. On the left, we plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Eq.~\ref{eq:high_alpha}.  On the right, we colour-code stars according to their [Mg/H] abundance.} \label{fig:subgiants}
\end{figure*}

@@

Given the challenge of {\it a priori} yield predictions, recent work aims to empirically characterize elemental production \dbnote{does it make sense to frame this as only recent work? I just don't know but I feel like there are very early examples which at least use observations as a guide for yields, like maybe even B2FH. or garnett1997}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and the metallicity dependence thereof.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.
\dbadd{Additionally the series of papers including \citet{eitner+2020} includes some relevant work.}

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in \add{REF 3.1} that the characteristic delay times of C production should be $\sim$ few hundred Myr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon...}

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (i.e., ``gravitational settling''; e.g. \dbadd{CITE}).
When stars evolve off main sequence, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
When a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}.

@@

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4, Gaia-ESO DR5.1, and LAMOST with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
\dbnote{Galah DR4 fixes the issues we were seeing and updates how they process C.}
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~7? and discussion in section \dbadd{5.2}).

@@

Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). We retain their prescription for O, Fe, an yields, scaled down to the normalization recommended by \citet{david_fe} based on .....
Table~\ref{tab:fiducial_mod} summarizes these choices. As required by \VICE, these yields quantify the mass of some element that is newly produced (i.e. net yields as opposed to gross yields). The SN yields are in units of the proginitors initial mass, while the AGB yields are defined relative to the star's zero-age main sequence mass.

\dbadd{paragraph about yields in general}
A yield quantifies the new production of a chemical species. For a given star, we define the {\it net fractional yield} to be the mean change in the chemical abundance from the birth to ejected mass, multiplied by the ejected fraction of a stars mass, i.e.
\begin{equation} 
   Y_{\rm C}^{\rm AGB}(Z) = 
    \delta Z_X \frac{M_{\rm ej}}{M}
\end{equation}
Note that yields may become negative if more of the initial chemical abundance of X is transformed into other elements than other elements transformed into X.
In general, we describe yields by their population-averaged forms, the amount of newly produced X per unit mass of star formation (from supernovae and stellar winds).
\begin{equation} 
   y_{X}(Z) = 
    \frac{M_X - M_{\rm X, ini}}{M_{\rm ssp}}
\end{equation}

In \VICE, massive star nucleosynthesis is approximated as occurring instantaneously after a star formation event. Under this approximation, the yield is simply a constant  of proportionality between the metal production rate and the SFR:
\begin{equation}
    \dot{M}_X^{\rm CC} = y_{X}^{\rm cc}\, \dot{M}_\star
\end{equation}
The population-averaged yield, $y_{X}^{\rm cc}$ can be expressed as
\begin{equation}
y_{X}^{\rm CC} = \frac{
\int_8^u (E(m) m_X + w_X - Z_X(m - m_{\rm rem}) )\frac{dN}{dM}\ dM
}{
\int_l^u m\,\frac{dN}{dM}\ dm
}
\end{equation}
where $m_X$ and $w_X$ are the mass of X ejected in a CCSN explosion and through stellar winds respectively, from a star of initial mass $m$. $E((m)$ is a function describing the fraction of stars of mass $m$ that successfully explode (se discussion in \S \dbadd{3.2} below). $m_{\rm rem}$ is the mass of the remnant left behind, $dN/dM$ is the IMF, for which we assume \dbadd{cite}. $l$ and $u$ are the lower and upper mass limits of star formation. $8\,\Mo$ is the minimum mass of a CCSN progenitor. In section \S 3.2 below, we compute the values of $\Ycc$ using \VICE's `vice.yields.ccsne.fractional` function and discuss theoretical expectations of C yields based on massive star evolutionary models.

The population averaged SNe Ia yield, $y_{X}^{\rm ia}$, is given by,
\begin{equation}
    y_{X}^{\rm Ia} = m_{x}^{\rm Ia} \frac{N_{\rm Ia}}{m_\star}
\end{equation}
where $m_{X}^{\rm Ia}$ is the average mass of X produced in a single SN Ia event, and $N_{\rm Ia} / M_\star$ is the number of such events per unit mass of star formation. The latter can be expressed as an integral over the SN Ia delay-time-distribution (DTD), which describes the rate at which stellar populations produce X through SN Ia in \VICE. Following \citet{james+23}, we use a $t^{-1.1}$ single power-law with a minimum delay-time of 150 Myr based on comparisons of the SN IA rate as a function of redshift and the cosmic SFR (e.g. \citealt{maoz+12}).

The net yield from a single AGB star is given by \begin{equation} \label{eq:net-yield}
   Y_{\rm C}^{\rm AGB}(Z) = 
    \delta Z_X \frac{M_{\rm ej}}{M}
\end{equation}
where $M_{\rm ej}$ is the mass of the ejected envelope, and $\delta Z_X$ is the average change in the abundance of X in the envelope over the lifetime of the star. Models of AGB star nucleosynthesis report values of $Y_X^{\rm AGB}$ sampled on a grid of initial masses and metallicities (see discussion in \S \dbadd{3.1}), and \VICE\ accepts these tables directly as input. The population-averaged AGB star yield is then given by 
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm agb}(Z,\tau) = 
    \frac{
    \int_{l}^{M_{\rm max}^{\rm AGB}} 
    \y_{\rm X}^{\rm proc}(M, Z) M\ 
    \frac{dN}{dM}\ dM
}
{
    \int_{l}^{u}\ M\ \frac{dN}{dM}\ dM
}
\end{equation}
where $m_{\rm min}$ is the minimum and maximum initial mass of stars that become AGBs. While $m_{\rm min}$ theoretically extends as low as $0.1 \Mo$, we interpolate yields to zero past $0.5 \Mo$ as these stars live longer than the disk and no yield grid extends this low. In Section~\ref{sec:agb},  we discuss theoretical expectations for C production in AGB stars and our adopted tables of $\Ycagb$ from the literature.
\dbadd{specify exactly how we change the yield interpolation for AGB stars}

\begin{table}
	\centering
    \caption[]{Solar abundance scale and fiducial yields (in units of stellar population birth mass). See section \ref{sec:agb} for the definition of \fruity. The solar abundance scale is \citet{magg+22} + 0.04. }
	\label{tab:fiducial_mod}

	\begin{tabular}{l l l l l}
		\hline
        Element & $Z_{X,\,\sun}$ & $y_X^{\rm cc}$ & $\y_X^{\rm agb}$ & $y_X^{\rm ia}$  \\
		\hline
        C & 0.00339 & Eq.~\ref{eq:y_cc} & $\cfactor\times$\fruity &  0 \\
        O & 0.00733 & 7.12e-3 & 0 & 0 \\
        Mg & 0.000671 & 6.52-4 & 0 & 0 \\
        Fe & 0.00137 & 4.72e-4 & 0 & 6.69e-4 \\
        N &0.00104 & 4.001e-4 & 0.0009$M\left(\frac{Z}{Z_\odot}\right)$ & 0\\
        % C & 0.00339 & Eq.~\ref{eq:zeta} & $\cfactor\times$\fruity &  0 \\
        % O & 0.00733 & 0.015 & 0 & 0 \\
        % Mg & 0.000671 & 0.00185 & 0 & 0 \\
        % Fe & 0.00137 & 0.0012 & 0 & 0.00214 \\
        % N &0.00104 & 0.00072 & 0.0009$M\left(\frac{Z}{Z_\odot}\right)$ & 0\\
		\hline
	\end{tabular}
\end{table}

@@

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{\dbnote{replace y label with stellar C yield, draw actual interpolations...}. The net fractional \agb\ C yield  plotted as a function of initial stellar mass $M$ and colour-coded according to metallicity. The black dashed line shows $\y=0$ for reference. Each panel represents yields from one of four \agb\ models: \fruity{}, \aton{}, \monash{}, \nugrid{} (see sections \ref{sec:agb}) }

@@

    \caption[]{\textbf{Left} \dbnote{ylabel: integrated carbon AGB yield} The (IMF-weighted) \agb\ C yield $\Ycagb$ as a function of metallicity for each of the \agb\ yield models. ($\Ycagb$ is the net mass of C produced by \agb\ stars per unit mass of star formation, after 10\,Gyr and assuming a \citealt{kroupa01} IMF.)
    \textbf{Right} cumulative C return as a function of age for a solar metallicity single stellar population. The dashed black line shows the delay time distribution of type Ia supernovae ($\propto t^{-1.1}$) for comparison. The minimum of \aton{} is at $y( 0.30\,{\rm Gyr})/y(t_{\rm end})=-7.81$ and is large as the net yield is close to zero.
}

@@

An Asymptotic Giant Branch (AGB) star is a low mass star during its last nuclear (He shell) burning phase which enriches the interstellar medium (ISM) through mass loss during thermal pulses.(see e.g. \citealt{PR2023}). 
In this work, we explore four different stellar AGB yield tables from the literature which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Table~\ref{tab:agb} notes the masses and metallicities for each grid of yields.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. \dbadd{the selection of tables is the same as CITE with the exception that ...}
For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \Ycagb \rightarrow \alpha_{\rm agb}\ \Ycagb.
\end{equation}
We use the \fruity\ table, with $\alpha_{\rm agb}=\cfactor$, as the fiducial \agb\ yield.

Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models.
\dbadd{there is good agreement} Most models agree on general yield trends in mass and metallicity.
C production peak between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly.

In order to integrate yields across the metallicity and mass range of interest, we interpolate yield tables linearly in both $Z$ and mass. Below the last sampled mass, $Y_{\rm C}^{\rm AGB}$ is linearly interpolated to 0 at $0.5\Mo$. The choice of yields for stars of masses $\Mo \lesssim 1.5$ produces about a 20\% uncertainty in the slope of the DTD past 3\,Gyr \dbstrike{but these old populations account for a small fraction of the total C}. \strike{More complete grids in mass and metallicity would be needed to more accurately compare GCE predictions between AGB evolution models. }

The left panel of Fig.~\ref{fig:agb-ssp} shows IMF-averaged C yields for each \agb\ model as a function of metallicity.
The IMF-averaged yields differ in normalization and metallicity dependence.  
The normalizations span a factor of \about{2}. (e.g. between $6\times 10^{-4}$ and $12 \times 10^{-4}$ at $Z=0.33\,\Zo$).
Both \nugrid{} and \aton{} predict non-monotonic C yields with $Z$, but it is unclear if this prediction is significant given the uncertainties associated with any one model.
To characterize the metallicity dependence of each table, we fit the yield to a linear model in $\log Z$
\begin{equation}\label{eq:agb_z_approx}
    y_{\rm C}^{\rm AGB}(Z) = y_0 + \zagb \log(Z / \Zo).
\end{equation}
Table \ref{tab:agb} contains the fit parameters $y_0$ and $\zagb$ for each AGB table.
The values of $\zagb$ span a factor of $\sim{3}$ between models.
These variations are due to different choices of reaction rates, convection treatments, and mass-loss rates \citep{ventura+15} \dbadd{REF}, but can be qualitatively understood by comparing the strength of third dredge up (TDU) and hot bottom burning (HBB) (see discussion below and \JJ).

The right panel of Fig.~\ref{fig:agb-ssp} shows the total production of C by \agb\ stars by a single stellar population (SSP; cluster of stars born at the same time from the same material) as a function of age at solar metallicity. 
As the mass range $2\,\Mo\lesssim M \lesssim 4\,\Mo$ is predicted to be most important for C production, half the yield is produced before \about{1}\,Gyr, similar to Fe production by \ia. 
\monash{} weight C production more heavily towards high-mass \agb\ stars, resulting in shorter delay times, whereas the \fruity\ and \aton\ models predict a slightly longer timescale of \about{1}\,Gyr. In any case, little to no C is produced more than 2\,Gyr after a star formation event. Fe production, in contrast, continues steadily for 10\,Gyr.

In AGB stars, {third dredge up} (TDU) and {hot bottom burning} (HBB) are two processes driving C evolution in \agb\ stars with competing effects.
TDU accompanies thermal pulses, where material from the CO core is mixed with the envelope, increasing surface C abundances 
later released to the ISM \citep{KL14}.  
HBB\ is the activation of the CNO cycle at the bottom of the convective envelope \dbadd{with a net effect of ...}.
TDU increases C yields and HBB converts C into N. When both processes are active, highly efficient N production ensues (see discussion in e.g. \citealt{james+23, ventura+13}). 
%Because the $^{14}$N proton capture is the slowest component of the CNO cycle, the CNO cycle converts nearly all \C[12] into $^{14}$N \citep{solar-fusion}.

@@

\section{Monte-Carlo Markov Chain inference}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,
% referee, % this line changes to double-spaced 1 column
usenatbib]{mnras}

@@

\newcommand{\fruity}{\texttt{\hyperlink{fruity}{Fruity}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{1.95}
\newcommand{\nsubgiants}{14,069}

% displays fractions within fractions well
\newcommand{\ddfrac}[2]{\ensuremath{%
    \frac{\displaystyle{#1}}{\displaystyle{#2}}
}}

@@

\newcommand{\Var}{\operatorname{Var}}

% other
\newcommand{\fmeas}{20\%}

@@

\newcommand{\Yct}{{y_{\rm C}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\ycagb}{\y_{\rm C}^{\rm AGB}}
\newcommand{\aagb}{\alpha_{\rm C}^{\rm AGB}}
\newcommand{\zagb}{\zeta_{\rm C}^{\rm AGB}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}
\newcommand{\noneqagb}{\nu_{\rm C}^{\rm AGB}}
 \newcommand{\yb}{\ensuremath{\rotatebox[origin=B,y=0.5ex]{180}{y}}}
\newcommand{\y}{p}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}

\newcommand{\zetao}{\zeta_0}
\newcommand{\zetai}{\zeta_{1}}
\newcommand{\zetaii}{\zeta_{2}}
\newcommand{\Zl}{x_{\rm low}}

@@

% Abstract of the paper
\begin{abstract}
% context
The astrophysical origin of carbon remains largely uncertain---different studies propose that either asymptotic giant branch (AGB) stars or core collapse supernovae (CCSNe) may be the dominant process.
% 
We aim to constrain the stellar yields of C through multi-zone Galactic chemical evolution models by comparing predictions with \apogee\ subgiants abundances.
% 
We show \caah\ trend represents the equilibrium abundances of C and Mg, deriving the total C yield as a function of metallicity. 
Instead, the \caafe\ trend (at a given metallicity) is an empirical estimate of the delay time distribution of C.
We infer that AGB stars produce between 10 and 40\% of C. Additionally, we show that all AGB models produce C too quickly, and the low-mass AGB C yield is likely underestimated. 
% misc
Our models are relatively independent of uniform scaling of yields and outflows, and alternate star formation histories. 
However, the stars which contribute to \agb\ C production and the \ia\ delay time distribution of Fe contribute uncertainties to our conclusions. 
% gas phase
While reliable gas-phase and low-metallicity measurements of C are challenging, we find that our model and a single-zone model with our recommended yields replicate the broad trends of \caah{} across different environments and metallicities.

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio against [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. On the left, we plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Eq.~\ref{eq:high_alpha}.  On the right, we colour-code stars according to their [Mg/H] abundance. The median subgiant trends serve as our emperical benchmark. } \label{fig:subgiants}
\end{figure*}

@@

Given the challenge of theoretical yield predictions, recent work aims to empirically characterize elemental production \dbnote{does it make sense to frame this as only recent work? I just don't know but I feel like there are very early examples which at least use observations as a guide for yields, like maybe even B2FH. or garnett1997}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and the metallicity dependence thereof.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.
\dbadd{Additionally the series of papers including \citet{eitner+2020} includes some relevant work.}

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their Fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in section~\ref{sec:agb} that the characteristic delay times of C production should be $\sim$ 1 Gyr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon...}

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (\textit{gravitational settling}; e.g. \citealt{turncotte+98}).
When stars evolve off main sequence, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
When a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}.
\dbnote{Why exactly do we use subgiants? Consistent stellar parameters, subgiant spectra are easier than white dwarfs? only a $\sim 0.05$ dex correction}

@@

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4, Gaia-ESO DR5.1, and LAMOST with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~7? and discussion in section \dbadd{5.2}).

@@

A yield quantifies the new production of a chemical species. For a given star, we define the {\it net fractional yield of $X$} as the fraction of a stars initial mass synthesized into new $X$. Specifically, the net fractional yield of element $X$ is
\begin{equation} 
   \y_{\rm X}(M, Z) = 
    \frac{m_{X, \rm ej}- Z_{X,\rm  ini}\,m_{\rm ej}}{m}
\end{equation}
where $Z_X$ is the initial abundance of $X$, $m_{\rm ej}$ is the ejected mass, $m_{X, \rm ej}$ is the ejected mass of $X$, and $m$ is the initial mass of the star.
Note that yields may become negative if more X is transformed into other elements than is produced ($Z_{X,\rm ej} < Z_{X,\rm ini}$).
We also describe yields by their population-averaged quantities, the amount of newly produced X per unit mass of star formation, 
\begin{equation} 
   y_{X}(Z) = 
    \frac{M_X - r\,Z_{X, \rm ini}\,M_{\rm ssp}}{M_{\rm ssp}},
\end{equation}
where $M_X$ is the total expelled mass of element X, $M_{\rm ssp}$ is the birth mass of the single stellar population (SSP\footnote{An SSP is a cluster of stars born at the same time from the same material.}), and $r$ is the return fraction (fraction of the SSP mass returned to the interstellar medium).

Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). We use the O, Mg, and Fe yields recommended by \citet{david_fe} based on measurements of the mean CCSN Fe yield \citep{rodriguez+23} and APOGEE chemical abundances.
Our N yield is scaled from \citep{james+23}.
Table~\ref{tab:fiducial_mod} summarizes these choices. 
We assume a \citet{kroupa01} initial mass function.

\begin{table}
	\centering
    \caption[]{Solar abundance scale and fiducial yields (in units of stellar population birth mass). See section \ref{sec:agb} for the definition of \fruity. The solar abundance scale is \citet{magg+22} + 0.04 dex suggested by \citet{david_fe}.
    \dbnote{any tips to make this table a little bit smaller?}
    }
	\label{tab:fiducial_mod}

	\begin{tabular}{l l l l l}
		\hline
         & $Z_{X,\,\sun}$ & $y_X^{\rm cc}$ & $\y_X^{\rm agb}$ & $y_X^{\rm ia}$  \\
		\hline
        C & $33.9\times10^{-4}$ & Eq.~\ref{eq:y_cc} & $\cfactor\times$\fruity &  0 \\
        O & $73.3\times10^{-4}$ & $71.3\times10^{-4}$ & 0 & 0 \\
        Mg & $6.71\times10^{-4}$ & $6.52\times 10^{-4}$ & 0 & 0 \\
        Fe & $13.7\times10^{-4}$ & $4.73\times10^{-4}$ & 0 & $7.82\times10^{-4}$ \\
        N & $10.4\times10^{-4}$ & $5\times10^{-4}$ & $5\times10^{-4}M\left(\frac{Z}{Z_\odot}\right)$ & 0\\
		\hline
	\end{tabular}
\end{table}

@@

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{The net fractional \agb\ C yield  plotted as a function of initial stellar mass and colour-coded according to metallicity. The black dashed line shows $\Ycagb=0$ for reference. Each panel represents yields from one of four \agb\ models: \fruity{}, \aton{}, \monash{}, \nugrid{} (see sections \ref{sec:agb}) }.

@@

    \caption[]{\textbf{Left} The integrated \agb\ C yield $\Ycagb$ as a function of metallicity for each of the \agb\ yield tables.
    \textbf{Right} cumulative C return fraction as a function of age for a solar metallicity single stellar population. The dashed black line shows the cumulative return fraction of type Ia supernovae ($\propto t^{-1.1}$) for comparison. The minimum of \aton{} is at $y( 0.30\,{\rm Gyr})/y(t_{\rm end})=-7.81$ and is large as the net yield is close to zero.
}

@@

An Asymptotic Giant Branch (AGB) star is a low mass star undergoing He shell burning, the last nuclear phase (see e.g. \citealt{PR2023}). AGB stars enrich the interstellar medium (ISM) through mass expelled during thermal pulses. 
In this work, we explore four different stellar AGB yield tables from the literature which provide well-sampled grids in mass and metallicity. We refer to the tables as 
\begin{description}
    \item \hypertarget{fruity}{\texttt{FRUITY}}: \citet{cristallo+11, cristallo+15}
    \item \hypertarget{aton}{\texttt{ATON}}: \citet{ventura+13,ventura+14,ventura+18, ventura+20}
    \item \hypertarget{monash}{\texttt{Monash}}: \citet{KL16, karakas+18}
    \item \hypertarget{nugrid}{\texttt{NuGrid}}: \citet{pignatari+16, ritter+18, battino+19, battino+21}
\end{description}
Each AGB yield table reports yields for stellar models on a grid of masses and metallicities as reported in Table~\ref{tab:agb}.
We use the same tables as \citet{james+23}, except that we have swapped \citet{karakas10} for \citet{pignatari+16}. 
For our models to better match observations, we uniformly amplify the yield tables according to
\begin{equation} \label{eq:alpha}
        \ycagb(M, Z) \rightarrow \alpha_{\rm agb}\ \ycagb(M, Z).
\end{equation}
We use the \fruity\ table, with $\alpha_{\rm agb}=\cfactor$, as the fiducial \agb\ yield.
Fig.~\ref{fig:y_agb} compares the C yields for these four AGB models.
Most models agree on general yield trends in mass and metallicity.
C production peak between masses of $\sim$ 2--4 $\Mo$. As metallicity increases, the net yield decreases and the mass of peak C production increases slightly.

In order to integrate AGB yields with \VICE, we linearly interpolate yield tables in both mass and metallicity. Below the smallest mass in a table, $\ycagb$ is linearly extrapolated to 0 at $1\Mo$. The population-averaged AGB star yield is calculated with 
\begin{equation} \label{eq:imf-agb-yield}
    y_{X}^{\rm agb}(Z,\tau) = 
    \frac{1}{A}\int^{u_{\rm AGB}}_{M_\textrm{post-AGB}(\tau)} 
    \y_{\rm X}(M, Z)\ 
    \frac{dN}{dM}\ M\ dM
\end{equation}
are the minimum and maximum mass of star formation, $u_{\rm AGB}=8\Mo$ is the maximum mass of an AGB star, $M_\textrm{post-AGB}(\tau)$ is the mass of stars which complete AGB evolution at age $\tau$, and $dN/dM$ is the initial mass function. $A = \int_l^u M \frac{dN}{dM} dm$ is the IMF normalization coefficient and  $l=0.08\,\Mo$ and $u = 100\,\Mo$. We assume a \citet{larson74} mass-lifetime-relationship and assume that the total lifetime of stars is 1.1 times the MS lifetime.

The left panel of Fig.~\ref{fig:agb-ssp} shows IMF-averaged C yields for each \agb\ model as a function of metallicity.
The IMF-averaged yields differ in normalization and metallicity dependence.  
The normalizations span a factor of \about{2}. (e.g. between $6\times 10^{-4}$ and $12 \times 10^{-4}$ at $Z=0.33\,\Zo$).
Both \nugrid{} and \aton{} predict non-monotonic C yields with $Z$, but it is unclear if this prediction is significant given the uncertainties associated with AGB modelling.
To characterize the metallicity dependence of each table, we fit the yield to a linear model in $\log Z$
\begin{equation}\label{eq:agb_z_approx}
    y_{\rm C}^{\rm AGB}(Z) = \zetao + \zetai \log(Z / \Zo).
\end{equation}
Table \ref{tab:agb} contains the fit parameters $\bar y_{\rm C, 0}^{\rm AGB}$ and $\zagb$ for each AGB table.
The values of $\zagb$ span a factor of $\sim{3}$ between models.
These variations are due to different choices of reaction rates, convection treatments, and mass-loss rates (see discussion below and \JJ).

The right panel of Fig.~\ref{fig:agb-ssp} shows the total relative production of C by \agb\ stars by a solar metallicity SSP as a function of age.
As the mass range $2\,\Mo\lesssim M \lesssim 4\,\Mo$ is predicted to be most important for C production, half the yield is produced before \about{1}\,Gyr, similar to Fe production by \ia. 
\monash{} weights C production more heavily towards high-mass \agb\ stars, resulting in shorter delay times, whereas the \fruity\ model predict a slightly longer timescale of \about{1}\,Gyr. In any case, little to no C is produced more than 2\,Gyr after a star formation event. Fe production, in contrast, continues steadily for 10\,Gyr.

In AGB stars, {third dredge up} (TDU) and {hot bottom burning} (HBB) are two competing processes driving C evolution.
TDU accompanies thermal pulses, where material from the CO core is mixed with the envelope, increasing surface C abundances 
later released to the ISM \citep{KL14}.  
HBB\ is the activation of the CNO cycle at the bottom of the convective envelope.
TDU increases C yields and HBB converts C into N. \dbstrike{When both processes are active, highly efficient N production ensues (see discussion in e.g. {james+23, ventura+13}).} \dbnote{move to next paragraph?}.
%Because the $^{14}$N proton capture is the slowest component of the CNO cycle, the CNO cycle converts nearly all \C[12] into $^{14}$N \citep{solar-fusion}.

@@

\section{Model parameters and details}

\begin{table}
	\centering
    \caption[]{Description of the models presented in this paper.}
	\label{tab:model_parameters}
